---
title: "Robust GFP Library Transfer Counts"
format: html
editor: visual
---

## Purpose

This notebook replaces the exploratory GFP library transfer-count analysis with a reproducible workflow that:

-   validates all required transfer-count inputs up front,
-   standardizes the eight GFP library conditions into one schema,
-   exports cleaned tables under `results/01_gfp_library_transfer_counts/`,
-   regenerates the main heatmap, kinetics, and viability plots.

## Setup

```{r}
resolve_project_root <- function() {
  rstudio_path <- ""
  if (requireNamespace("rstudioapi", quietly = TRUE)) {
    rstudio_path <- tryCatch(rstudioapi::getSourceEditorContext()$path, error = function(e) "")
    if (!nzchar(rstudio_path)) {
      rstudio_path <- tryCatch(rstudioapi::getActiveDocumentContext()$path, error = function(e) "")
    }
  }

  search_roots <- unique(c(
    Sys.getenv("CAS12A_NOTEBOOK_PROJECT_ROOT", unset = ""),
    tryCatch(knitr::current_input(dir = TRUE), error = function(e) ""),
    rstudio_path,
    getwd(),
    normalizePath(file.path(getwd(), ".."), mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), mustWork = FALSE)
  ))
  search_roots <- search_roots[nzchar(search_roots)]

  for (root in search_roots) {
    current <- normalizePath(root, mustWork = FALSE)
    if (!dir.exists(current)) {
      current <- dirname(current)
    }

    repeat {
      required_paths <- c(
        file.path(current, "_quarto.yml"),
        file.path(current, "config", "paths.R"),
        file.path(current, "R", "common.R")
      )

      if (all(file.exists(required_paths))) {
        return(current)
      }

      parent <- dirname(current)
      if (identical(parent, current)) {
        break
      }

      current <- parent
    }
  }

  stop(
    paste(
      "Could not locate the notebook project root.",
      "Save the notebook and either run it from the project root,",
      "or set Sys.setenv(CAS12A_NOTEBOOK_PROJECT_ROOT = '/path/to/New project')."
    ),
    call. = FALSE
  )
}

project_root <- resolve_project_root()

source(file.path(project_root, "config", "paths.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "common.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "gfp_library.R"), local = knitr::knit_global())

load_paper_packages()

paths <- get_analysis_paths(
  list(results_dir = normalizePath(file.path(project_root, "results"), mustWork = FALSE))
)

output_dir <- analysis_output_dir(paths, "01_gfp_library_transfer_counts")
tables_dir <- ensure_dir(file.path(output_dir, "tables"))
plots_dir <- ensure_dir(file.path(output_dir, "plots"))

control_well <- "A09"
selected_heatmap_timepoint <- "6_gen"
```

## Inputs

```{r}
gfp_transfer <- load_all_gfp_transfer_counts(paths)

manifest <- gfp_transfer$manifest
all_replicates <- gfp_transfer$all_replicates
condition_averages <- prepare_gfp_condition_averages(all_replicates)
all_replicates_with_viability <- compute_gfp_viability(all_replicates, control_well = control_well)
first_timepoint_summary <- summarize_first_timepoint(
  all_replicates_with_viability,
  timepoint_name = "scraped",
  control_well = control_well
)

write_csv_checked(manifest, file.path(tables_dir, "manifest.csv"))
write_csv_checked(all_replicates, file.path(tables_dir, "standardized_all_replicates.csv"))
write_csv_checked(condition_averages, file.path(tables_dir, "condition_averages.csv"))
write_csv_checked(first_timepoint_summary, file.path(tables_dir, "first_timepoint_summary.csv"))

manifest
```

## Heatmaps

```{r}
enas_heatmaps <- plot_gfp_heatmaps(
  condition_averages %>% dplyr::filter(cas_variant == "enAsCas12a"),
  plot_title = "GFP library: enAsCas12a",
  fill_high = "#E69F00",
  control_well = control_well
)

fn_heatmaps <- plot_gfp_heatmaps(
  condition_averages %>% dplyr::filter(cas_variant == "FnCas12a"),
  plot_title = "GFP library: FnCas12a",
  fill_high = "#009E73",
  control_well = control_well
)

stopifnot(selected_heatmap_timepoint %in% names(enas_heatmaps))
stopifnot(selected_heatmap_timepoint %in% names(fn_heatmaps))

safe_ggsave(
  enas_heatmaps[[selected_heatmap_timepoint]],
  file.path(plots_dir, "enAs_gfp_heat_6gen.pdf"),
  width = 10,
  height = 6
)

safe_ggsave(
  fn_heatmaps[[selected_heatmap_timepoint]],
  file.path(plots_dir, "fn_gfp_heat_6gen.pdf"),
  width = 10,
  height = 6
)

enas_heatmaps[[selected_heatmap_timepoint]]
```

```{r}
fn_heatmaps[[selected_heatmap_timepoint]]
```

## Kinetics

```{r}
enas_kinetics_plot <- plot_gfp_kinetics(all_replicates, "enAsCas12a")
fn_kinetics_plot <- plot_gfp_kinetics(all_replicates, "FnCas12a")

safe_ggsave(
  enas_kinetics_plot,
  file.path(plots_dir, "enas_kinetics_all_guides.png"),
  width = 10,
  height = 6
)

safe_ggsave(
  fn_kinetics_plot,
  file.path(plots_dir, "fn_kinetics_all_guides.png"),
  width = 10,
  height = 6
)

enas_kinetics_plot
```

## Viability

```{r}
plot_first_timepoint_viability <- function(data, cas_variant_name) {
  plot_data <- data %>%
    dplyr::filter(cas_variant == cas_variant_name) %>%
    dplyr::mutate(donor_recruitment = label_recruitment(donor_recruitment))

  ggplot(plot_data, aes(x = viability, y = efficiency_avg)) +
    geom_point(aes(color = well_position), size = 3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    ggpmisc::stat_poly_eq(
      aes(label = after_stat(rr.label)),
      formula = y ~ x,
      parse = TRUE,
      label.x.npc = "left",
      label.y.npc = "top"
    ) +
    facet_wrap(
      ~ donor_recruitment * promoter,
      labeller = labeller(donor_recruitment = c("LexA-FHA" = "LexA-FHA", "No" = ""))
    ) +
    labs(
      x = "Viability (relative to A09 control)",
      y = "Average efficiency (%)",
      title = paste("First timepoint viability:", cas_variant_name)
    ) +
    publication_theme(legend_position = "none", aspect_ratio = 1)
}

viability_enas_plot <- plot_first_timepoint_viability(first_timepoint_summary, "enAsCas12a")
viability_fn_plot <- plot_first_timepoint_viability(first_timepoint_summary, "FnCas12a")

safe_ggsave(
  viability_enas_plot,
  file.path(plots_dir, "viability_efficiency_enas_gfp_lib.png"),
  width = 6,
  height = 6
)

safe_ggsave(
  viability_fn_plot,
  file.path(plots_dir, "viability_efficiency_fn_gfp_lib.png"),
  width = 6,
  height = 6
)

viability_enas_plot
```

```{r}
variant_boxplot <- first_timepoint_summary %>%
  dplyr::mutate(donor_recruitment = label_recruitment(donor_recruitment)) %>%
  ggplot(aes(cas_variant, viability, fill = cas_variant)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, size = 1.5, aes(color = well_position)) +
  scale_fill_manual(values = c("enAsCas12a" = "#E69F00", "FnCas12a" = "#009E73")) +
  facet_wrap(
    ~ donor_recruitment * promoter,
    labeller = labeller(donor_recruitment = c("LexA-FHA" = "LexA-FHA", "No" = ""))
  ) +
  labs(x = "Cas variant", y = "Viability (relative to A09 control)") +
  publication_theme(legend_position = "none", aspect_ratio = 1)

safe_ggsave(
  variant_boxplot,
  file.path(plots_dir, "viability_boxplot_gfp_lib.pdf"),
  width = 6,
  height = 6
)

variant_boxplot
```
