---
title: "Robust Cas Variant Overview"
format: html
editor: visual
---

## Purpose

This notebook replaces the mixed GFP ON/OFF Cas-variant notebook with a structured workflow that rebuilds processed tables from raw inputs and exports the main panels under `results/04_cas_variants_overview/`.

## Setup

```{r}
resolve_project_root <- function() {
  rstudio_path <- ""
  if (requireNamespace("rstudioapi", quietly = TRUE)) {
    rstudio_path <- tryCatch(rstudioapi::getSourceEditorContext()$path, error = function(e) "")
    if (!nzchar(rstudio_path)) {
      rstudio_path <- tryCatch(rstudioapi::getActiveDocumentContext()$path, error = function(e) "")
    }
  }

  search_roots <- unique(c(
    Sys.getenv("CAS12A_NOTEBOOK_PROJECT_ROOT", unset = ""),
    tryCatch(knitr::current_input(dir = TRUE), error = function(e) ""),
    rstudio_path,
    getwd(),
    normalizePath(file.path(getwd(), ".."), mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), mustWork = FALSE)
  ))
  search_roots <- search_roots[nzchar(search_roots)]

  for (root in search_roots) {
    current <- normalizePath(root, mustWork = FALSE)
    if (!dir.exists(current)) {
      current <- dirname(current)
    }

    repeat {
      required_paths <- c(
        file.path(current, "_quarto.yml"),
        file.path(current, "config", "paths.R"),
        file.path(current, "R", "common.R")
      )

      if (all(file.exists(required_paths))) {
        return(current)
      }

      parent <- dirname(current)
      if (identical(parent, current)) {
        break
      }

      current <- parent
    }
  }

  stop(
    paste(
      "Could not locate the notebook project root.",
      "Save the notebook and either run it from the project root,",
      "or set Sys.setenv(CAS12A_NOTEBOOK_PROJECT_ROOT = '/path/to/New project')."
    ),
    call. = FALSE
  )
}

project_root <- resolve_project_root()

source(file.path(project_root, "config", "paths.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "common.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "cas_variants.R"), local = knitr::knit_global())

load_paper_packages()

paths <- get_analysis_paths(
  list(results_dir = normalizePath(file.path(project_root, "results"), mustWork = FALSE))
)

output_dir <- analysis_output_dir(paths, "04_cas_variants_overview")
tables_dir <- ensure_dir(file.path(output_dir, "tables"))
plots_dir <- ensure_dir(file.path(output_dir, "plots"))
```

## Build processed tables

```{r}
gfp_on <- build_gfp_on_variants(paths)
gfp_off_list <- build_gfp_off_variants(paths)
direct_repeat_tables <- build_direct_repeat_final_tables(paths)
guide_promoter_tables <- build_guide_promoter_final_tables(paths)
gfp_off <- gfp_off_list$updated

write_csv_checked(gfp_on, file.path(tables_dir, "gfp_on_variants.csv"))
write_csv_checked(gfp_off, file.path(tables_dir, "gfp_off_variants.csv"))
write_csv_checked(direct_repeat_tables$with_colony_counts, file.path(tables_dir, "gfp_off_direct_repeat.csv"))
write_csv_checked(direct_repeat_tables$syn1, file.path(tables_dir, "gfp_off_direct_repeat_syn1.csv"))
write_csv_checked(guide_promoter_tables$base, file.path(tables_dir, "guide_promoter_data.csv"))
write_csv_checked(
  guide_promoter_tables$plate_counts,
  file.path(tables_dir, "guide_promoter_plate_counts_20240916.csv")
)
write_csv_checked(
  guide_promoter_tables$with_colony_counts,
  file.path(tables_dir, "guide_promoter_data_with_colony_counts.csv")
)

gfp_on
```

## Main GFP ON panels

```{r}
tef_on_plot <- plot_variant_timecourse(
  gfp_on,
  cas_promoter = "TEF",
  exclude = c("enAsCas12a_less_NLS", "FnCas12a(CO)", "LbCas12a(CO)"),
  facet = TRUE,
  nrows = 1
)

pgk1_on_plot <- plot_variant_timecourse(
  gfp_on,
  cas_promoter = "PGK1",
  exclude = c("FnCas12a(CO)", "LbCas12a(CO)"),
  facet = TRUE,
  nrows = 1
)

safe_ggsave(
  tef_on_plot,
  file.path(plots_dir, "tef_variants_on.pdf"),
  width = 7,
  height = 4
)

safe_ggsave(
  pgk1_on_plot,
  file.path(plots_dir, "pgk1_variants_on.pdf"),
  width = 7,
  height = 4
)

tef_on_plot
```

## Main GFP OFF panels

```{r}
tef_off_plot <- plot_variant_timecourse(
  gfp_off,
  cas_promoter = "TEF",
  exclude = c("FnCas12a(CO)", "LbCas12a(CO)"),
  facet = TRUE,
  nrows = 1
)

pgk1_off_plot <- plot_variant_timecourse(
  gfp_off,
  cas_promoter = "PGK1",
  exclude = c("enAsCas12a_less_NLS", "FnCas12a(CO)", "LbCas12a(CO)"),
  facet = TRUE,
  nrows = 1
)

safe_ggsave(
  tef_off_plot,
  file.path(plots_dir, "tef_variants_off.pdf"),
  width = 7,
  height = 4
)

safe_ggsave(
  pgk1_off_plot,
  file.path(plots_dir, "pgk1_variants_off.pdf"),
  width = 7,
  height = 4
)

tef_off_plot
```

## Direct repeat comparison

```{r}
direct_repeat_plot <- plot_direct_repeat_by_guide_promoter(
  direct_repeat_tables$with_colony_counts,
  cas_promoter = "TEF",
  exclude = c("enAsCas12a_less_NLS", "FnCas12a(CO)", "LbCas12a(CO)"),
  nrows = 1
)

safe_ggsave(
  direct_repeat_plot,
  file.path(plots_dir, "direct_repeat_tef.pdf"),
  width = 8,
  height = 4
)

direct_repeat_plot
```

## Codon-optimized supplementary panel

```{r}
co_data <- gfp_on %>%
  dplyr::filter(var %in% c("FnCas12a", "LbCas12a", "FnCas12a(CO)", "LbCas12a(CO)")) %>%
  dplyr::mutate(parent_var = dplyr::if_else(stringr::str_detect(var, "Fn"), "Fn", "Lb"))

co_tef_plot <- plot_codon_optimized_comparison(co_data, "TEF")
co_pgk1_plot <- plot_codon_optimized_comparison(co_data, "PGK1")

safe_ggsave(
  co_tef_plot,
  file.path(plots_dir, "codon_optimized_tef.pdf"),
  width = 7,
  height = 4
)

safe_ggsave(
  co_pgk1_plot,
  file.path(plots_dir, "codon_optimized_pgk1.pdf"),
  width = 7,
  height = 4
)

co_tef_plot
```

## Guide promoter exploration

```{r}
guide_promoter_on_plot <- plot_guide_promoter_timecourse(guide_promoter_tables$with_colony_counts, "ON")
guide_promoter_off_plot <- plot_guide_promoter_timecourse(guide_promoter_tables$with_colony_counts, "OFF")

safe_ggsave(
  guide_promoter_on_plot,
  file.path(plots_dir, "guide_promoter_on.pdf"),
  width = 8,
  height = 4
)

safe_ggsave(
  guide_promoter_off_plot,
  file.path(plots_dir, "guide_promoter_off.pdf"),
  width = 8,
  height = 4
)

guide_promoter_on_plot
```
