---
title: "Genome-wide Cas12a Design Rules"
format: html
editor: visual
---

## Purpose

This notebook rebuilds the genome-wide picked-colony editing screen into a shareable design-rules workflow. It integrates the original Tn5 coverage and normalized-variant logic with workspace guide annotations (`DeepCpf1`, SV hotspot context, and variant functional annotation) and exports figure-ready panels under `results/06_genome_wide_design_rules/`.

## Setup

```{r}
resolve_project_root <- function() {
  rstudio_path <- ""
  if (requireNamespace("rstudioapi", quietly = TRUE)) {
    rstudio_path <- tryCatch(rstudioapi::getSourceEditorContext()$path, error = function(e) "")
    if (!nzchar(rstudio_path)) {
      rstudio_path <- tryCatch(rstudioapi::getActiveDocumentContext()$path, error = function(e) "")
    }
  }

  search_roots <- unique(c(
    Sys.getenv("CAS12A_NOTEBOOK_PROJECT_ROOT", unset = ""),
    tryCatch(knitr::current_input(dir = TRUE), error = function(e) ""),
    rstudio_path,
    getwd(),
    normalizePath(file.path(getwd(), ".."), mustWork = FALSE),
    normalizePath(file.path(getwd(), "..", ".."), mustWork = FALSE)
  ))
  search_roots <- search_roots[nzchar(search_roots)]

  for (root in search_roots) {
    current <- normalizePath(root, mustWork = FALSE)
    if (!dir.exists(current)) {
      current <- dirname(current)
    }

    repeat {
      required_paths <- c(
        file.path(current, "_quarto.yml"),
        file.path(current, "config", "paths.R"),
        file.path(current, "R", "common.R")
      )

      if (all(file.exists(required_paths))) {
        return(current)
      }

      parent <- dirname(current)
      if (identical(parent, current)) {
        break
      }

      current <- parent
    }
  }

  stop(
    paste(
      "Could not locate the notebook project root.",
      "Save the notebook and either run it from the project root,",
      "or set Sys.setenv(CAS12A_NOTEBOOK_PROJECT_ROOT = '/path/to/New project')."
    ),
    call. = FALSE
  )
}

project_root <- resolve_project_root()

source(file.path(project_root, "config", "paths.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "common.R"), local = knitr::knit_global())
source(file.path(project_root, "R", "genome_wide_design_rules.R"), local = knitr::knit_global())

load_paper_packages(extra = c("broom", "forcats", "scales"))

paths <- get_analysis_paths(
  list(results_dir = normalizePath(file.path(project_root, "results"), mustWork = FALSE))
)

output_dir <- analysis_output_dir(paths, "06_genome_wide_design_rules")
tables_dir <- ensure_dir(file.path(output_dir, "tables"))
plots_dir <- ensure_dir(file.path(output_dir, "plots"))
```

## Build cleaned assay tables

```{r}
design_rule_objects <- build_genome_wide_design_rule_tables(project_root, paths)

colony_assays <- design_rule_objects$colony_assays
design_outcomes <- design_rule_objects$design_outcomes

stage_summary <- build_genome_wide_stage_summary(design_outcomes)
pam_summary <- summarise_edit_rates(design_outcomes, "pam") %>%
  dplyr::mutate(pam = factor(pam, levels = c("TTTC", "TTTA", "TTTG")))
pam_distance_summary <- summarise_edit_rates(design_outcomes, "pam_dist_bin") %>%
  dplyr::mutate(pam_dist_bin = factor(pam_dist_bin, levels = c("1-6", "7-11", "12-17", "18-23", "PAM")))
pam_heatmap_summary <- summarise_edit_rates(design_outcomes, c("pam", "pam_dist_bin")) %>%
  dplyr::filter(!is.na(pam_dist_bin), pam_dist_bin != ">23")
deepcpf1_summary <- summarise_edit_rates(design_outcomes, "deepcpf1_quartile")
pam_upstream_summary <- summarise_edit_rates(design_outcomes, "pam_upstream_nt") %>%
  dplyr::mutate(pam_upstream_nt = factor(pam_upstream_nt, levels = c("C", "A", "G", "T")))
full_pam_5mer_summary <- summarise_edit_rates(design_outcomes, "full_pam_5mer") %>%
  dplyr::arrange(full_pam_5mer)
ntttv_pattern_summary <- summarise_edit_rates(design_outcomes, "ntttv_pattern") %>%
  dplyr::arrange(dplyr::desc(total_guides), ntttv_pattern)
rule_count_summary <- summarise_edit_rates(design_outcomes, "preferred_rule_count") %>%
  dplyr::mutate(preferred_rule_count = factor(preferred_rule_count, levels = 0:3))
dr_perturbed_summary <- design_outcomes %>%
  dplyr::mutate(
    dr_perturbed_flag = dplyr::if_else(.data$dr_perturbed, "Perturbed", "Not perturbed"),
    dr_perturbed_flag = factor(dr_perturbed_flag, levels = c("Not perturbed", "Perturbed"))
  ) %>%
  summarise_edit_rates("dr_perturbed_flag")
dr_canonical_summary <- design_outcomes %>%
  dplyr::mutate(
    dr_canonical_flag = dplyr::if_else(.data$dr_canonical_structure, "Canonical", "Non-canonical"),
    dr_canonical_flag = factor(dr_canonical_flag, levels = c("Canonical", "Non-canonical"))
  ) %>%
  summarise_edit_rates("dr_canonical_flag")
dr_spacer_pair_summary <- design_outcomes %>%
  dplyr::mutate(
    dr_spacer_pair_flag = dplyr::if_else(.data$dr_spacer_pair_any, "Any DR-spacer pairing", "No DR-spacer pairing"),
    dr_spacer_pair_flag = factor(
      dr_spacer_pair_flag,
      levels = c("No DR-spacer pairing", "Any DR-spacer pairing")
    )
  ) %>%
  summarise_edit_rates("dr_spacer_pair_flag")
guide_gc_23_summary <- summarise_edit_rates(design_outcomes, "guide_gc_23_bin") %>%
  dplyr::mutate(guide_gc_23_bin = factor(guide_gc_23_bin, levels = c("<30%", "30-39%", "40-49%", "50-59%", ">=60%")))
guide_gc_20_summary <- summarise_edit_rates(design_outcomes, "guide_gc_20_bin") %>%
  dplyr::mutate(guide_gc_20_bin = factor(guide_gc_20_bin, levels = c("<30%", "30-39%", "40-49%", "50-59%", ">=60%")))
guide_gc_18_summary <- summarise_edit_rates(design_outcomes, "guide_gc_18_bin") %>%
  dplyr::mutate(guide_gc_18_bin = factor(guide_gc_18_bin, levels = c("<30%", "30-39%", "40-49%", "50-59%", ">=60%")))
donor_wt_gc_summary <- summarise_edit_rates(design_outcomes, "donor_wt_gc_bin") %>%
  dplyr::mutate(donor_wt_gc_bin = factor(donor_wt_gc_bin, levels = c("<30%", "30-39%", "40-49%", "50-59%", ">=60%")))
max_a_summary <- summarise_edit_rates(design_outcomes, "max_a") %>%
  dplyr::mutate(max_a = factor(max_a, levels = sort(unique(max_a))))
max_t_summary <- summarise_edit_rates(design_outcomes, "max_t") %>%
  dplyr::mutate(max_t = factor(max_t, levels = sort(unique(max_t))))
region_class_summary <- design_outcomes %>%
  dplyr::mutate(region_class = tidyr::replace_na(region_class, "unknown")) %>%
  summarise_edit_rates("region_class") %>%
  dplyr::mutate(region_class = factor(region_class, levels = c("coding", "promoter", "intergenic", "unknown")))
hotspot_summary <- design_outcomes %>%
  dplyr::mutate(
    hotspot_flag = dplyr::if_else(
      dplyr::coalesce(.data$in_sv_hotspot, FALSE),
      "SV hotspot",
      "Not hotspot"
    ),
    hotspot_flag = factor(hotspot_flag, levels = c("Not hotspot", "SV hotspot"))
  ) %>%
  summarise_edit_rates("hotspot_flag")
multi_colony_consistency_summary <- build_multi_colony_consistency_summary(design_outcomes)
design_rule_summary <- build_genome_wide_design_rule_summary(design_outcomes)

background_variant_tables <- build_genome_wide_background_variant_tables(
  design_rule_objects$files,
  colony_assays,
  qual_min = 20,
  af_min = 50,
  dp_min = 4
)

background_colony_summary <- background_variant_tables$colony_background_summary
background_edit_status_summary <- background_colony_summary %>%
  dplyr::mutate(target_edit_status = dplyr::if_else(.data$edited_colony, "Edited target", "Unedited target")) %>%
  dplyr::group_by(.data$target_edit_status) %>%
  dplyr::summarise(
    colony_count = dplyr::n(),
    mean_high_conf_non_target_count = mean(.data$high_conf_non_target_count, na.rm = TRUE),
    median_high_conf_non_target_count = median(.data$high_conf_non_target_count, na.rm = TRUE),
    frac_with_any_high_conf_non_target = mean(.data$high_conf_non_target_count > 0, na.rm = TRUE),
    mean_max_non_target_score_sv = mean(.data$max_non_target_score_sv, na.rm = TRUE),
    .groups = "drop"
  )

recurrent_background_variants <- background_variant_tables$recurrent_high_conf_variants %>%
  dplyr::filter(.data$sample_count >= 2)

weighted_model <- fit_genome_wide_weighted_model(design_outcomes)
weighted_model_table <- tidy_genome_wide_weighted_model(weighted_model)
extended_weighted_model <- fit_genome_wide_extended_weighted_model(design_outcomes)
extended_weighted_model_table <- tidy_genome_wide_extended_weighted_model(extended_weighted_model)

write_csv_checked(colony_assays, file.path(tables_dir, "colony_assays.csv"))
write_csv_checked(design_outcomes, file.path(tables_dir, "design_outcomes.csv"))
write_csv_checked(stage_summary, file.path(tables_dir, "stage_summary.csv"))
write_csv_checked(pam_summary, file.path(tables_dir, "pam_summary.csv"))
write_csv_checked(pam_distance_summary, file.path(tables_dir, "pam_distance_summary.csv"))
write_csv_checked(pam_heatmap_summary, file.path(tables_dir, "pam_distance_heatmap_summary.csv"))
write_csv_checked(deepcpf1_summary, file.path(tables_dir, "deepcpf1_summary.csv"))
write_csv_checked(pam_upstream_summary, file.path(tables_dir, "pam_upstream_summary.csv"))
write_csv_checked(full_pam_5mer_summary, file.path(tables_dir, "full_pam_5mer_summary.csv"))
write_csv_checked(ntttv_pattern_summary, file.path(tables_dir, "ntttv_pattern_summary.csv"))
write_csv_checked(rule_count_summary, file.path(tables_dir, "preferred_rule_count_summary.csv"))
write_csv_checked(dr_perturbed_summary, file.path(tables_dir, "dr_perturbed_summary.csv"))
write_csv_checked(dr_canonical_summary, file.path(tables_dir, "dr_canonical_summary.csv"))
write_csv_checked(dr_spacer_pair_summary, file.path(tables_dir, "dr_spacer_pair_summary.csv"))
write_csv_checked(guide_gc_23_summary, file.path(tables_dir, "guide_gc_23_summary.csv"))
write_csv_checked(guide_gc_20_summary, file.path(tables_dir, "guide_gc_20_summary.csv"))
write_csv_checked(guide_gc_18_summary, file.path(tables_dir, "guide_gc_18_summary.csv"))
write_csv_checked(donor_wt_gc_summary, file.path(tables_dir, "donor_wt_gc_summary.csv"))
write_csv_checked(max_a_summary, file.path(tables_dir, "max_a_summary.csv"))
write_csv_checked(max_t_summary, file.path(tables_dir, "max_t_summary.csv"))
write_csv_checked(region_class_summary, file.path(tables_dir, "region_class_summary.csv"))
write_csv_checked(hotspot_summary, file.path(tables_dir, "hotspot_summary.csv"))
write_csv_checked(multi_colony_consistency_summary, file.path(tables_dir, "multi_colony_consistency_summary.csv"))
write_csv_checked(design_rule_summary, file.path(tables_dir, "design_rule_summary.csv"))
write_csv_checked(background_variant_tables$all_variants, file.path(tables_dir, "all_variants_with_score_sv.csv"))
write_csv_checked(
  background_variant_tables$high_conf_non_target_variants,
  file.path(tables_dir, "high_conf_non_target_variants.csv")
)
write_csv_checked(background_colony_summary, file.path(tables_dir, "background_colony_summary.csv"))
write_csv_checked(background_edit_status_summary, file.path(tables_dir, "background_edit_status_summary.csv"))
write_csv_checked(recurrent_background_variants, file.path(tables_dir, "recurrent_background_variants.csv"))
write_csv_checked(weighted_model_table, file.path(tables_dir, "weighted_model_coefficients.csv"))
write_csv_checked(extended_weighted_model_table, file.path(tables_dir, "extended_weighted_model_coefficients.csv"))

tibble::tibble(
  metric = c(
    "Picked colonies",
    "Assayed colonies",
    "Edited colonies",
    "Unique picked designs",
    "Assayed designs",
    "Successful designs",
    "Direct-repeat sequence",
    "High-confidence non-target variants"
  ),
  value = c(
    nrow(colony_assays),
    sum(colony_assays$assayed_colony),
    sum(colony_assays$edited_colony),
    nrow(design_outcomes),
    sum(design_outcomes$assayed_design),
    sum(design_outcomes$edited_design),
    genome_wide_dr_sequence(),
    nrow(background_variant_tables$high_conf_non_target_variants)
  )
)
```

## Candidate main-figure panels

```{r}
pam_distance_heatmap <- plot_genome_wide_pam_distance_heatmap(pam_heatmap_summary)
deepcpf1_plot <- plot_genome_wide_rate_bars(
  deepcpf1_summary,
  x_col = "deepcpf1_quartile",
  x_label = "DeepCpf1 quartile",
  title = "Higher DeepCpf1 scores enrich successful designs"
)
pam_upstream_plot <- plot_genome_wide_rate_bars(
  pam_upstream_summary,
  x_col = "pam_upstream_nt",
  x_label = "5' PAM base (N in NTTTV)",
  title = "The upstream PAM base adds additional design signal"
)
rule_count_plot <- plot_genome_wide_rate_bars(
  rule_count_summary,
  x_col = "preferred_rule_count",
  x_label = "Preferred-rule count",
  title = "Simple rule bundles separate strong from weak designs"
)
multi_colony_consistency_plot <- plot_multi_colony_consistency(multi_colony_consistency_summary)
extended_model_plot <- plot_genome_wide_weighted_model(extended_weighted_model_table)

safe_ggsave(pam_distance_heatmap, file.path(plots_dir, "pam_distance_heatmap.pdf"), width = 6.5, height = 4.5)
safe_ggsave(deepcpf1_plot, file.path(plots_dir, "deepcpf1_quartiles.pdf"), width = 5.2, height = 4.8)
safe_ggsave(pam_upstream_plot, file.path(plots_dir, "pam_upstream_nt.pdf"), width = 5.2, height = 4.8)
safe_ggsave(rule_count_plot, file.path(plots_dir, "preferred_rule_count.pdf"), width = 5.2, height = 4.8)
safe_ggsave(multi_colony_consistency_plot, file.path(plots_dir, "multi_colony_consistency.pdf"), width = 5.5, height = 5)
safe_ggsave(extended_model_plot, file.path(plots_dir, "extended_weighted_model_odds_ratios.pdf"), width = 8.5, height = 5.5)

candidate_main_figure <- (
  (pam_distance_heatmap | deepcpf1_plot) /
  (pam_upstream_plot | multi_colony_consistency_plot)
) +
  patchwork::plot_annotation(tag_levels = "A")

safe_ggsave(
  candidate_main_figure,
  file.path(plots_dir, "candidate_main_figure_design_rules.pdf"),
  width = 11,
  height = 10
)

candidate_main_figure
```

## PAM 5' base and direct-repeat folding

```{r}
dr_perturbed_plot <- plot_genome_wide_rate_bars(
  dr_perturbed_summary,
  x_col = "dr_perturbed_flag",
  x_label = "DR folding call",
  title = "Perturbed DR folding alone was not a major failure mode"
)
dr_canonical_plot <- plot_genome_wide_rate_bars(
  dr_canonical_summary,
  x_col = "dr_canonical_flag",
  x_label = "DR structure class",
  title = "Canonical DR structures were not strongly enriched"
)
dr_spacer_pair_plot <- plot_genome_wide_rate_bars(
  dr_spacer_pair_summary,
  x_col = "dr_spacer_pair_flag",
  x_label = "DR-spacer pairing",
  title = "Any DR-spacer pairing had only a weak marginal effect"
)

safe_ggsave(dr_perturbed_plot, file.path(plots_dir, "dr_perturbed_summary.pdf"), width = 5.4, height = 4.8)
safe_ggsave(dr_canonical_plot, file.path(plots_dir, "dr_canonical_summary.pdf"), width = 5.4, height = 4.8)
safe_ggsave(dr_spacer_pair_plot, file.path(plots_dir, "dr_spacer_pair_summary.pdf"), width = 5.4, height = 4.8)

dr_support_figure <- (
  pam_upstream_plot | dr_perturbed_plot | dr_spacer_pair_plot
) +
  patchwork::plot_annotation(tag_levels = "A")

safe_ggsave(
  dr_support_figure,
  file.path(plots_dir, "pam_upstream_and_dr_folding.pdf"),
  width = 15,
  height = 4.8
)

dr_support_figure
```

```{r}
ntttv_pattern_summary %>%
  dplyr::mutate(
    design_success_rate = scales::percent(design_success_rate, accuracy = 0.1),
    colony_edit_rate = scales::percent(colony_edit_rate, accuracy = 0.1)
  ) %>%
  dplyr::select(ntttv_pattern, total_guides, design_success_rate, colony_edit_rate) %>%
  dplyr::slice_head(n = 12)
```

## GC dependencies across guide and donor

```{r}
guide_gc_23_plot <- plot_genome_wide_rate_bars(
  guide_gc_23_summary,
  x_col = "guide_gc_23_bin",
  x_label = "Guide GC (23 nt)",
  title = "Full spacer GC content"
)
guide_gc_20_plot <- plot_genome_wide_rate_bars(
  guide_gc_20_summary,
  x_col = "guide_gc_20_bin",
  x_label = "Guide GC (first 20 nt)",
  title = "Guide GC, first 20 nt"
)
guide_gc_18_plot <- plot_genome_wide_rate_bars(
  guide_gc_18_summary,
  x_col = "guide_gc_18_bin",
  x_label = "Guide GC (first 18 nt)",
  title = "Guide GC, first 18 nt"
)
donor_wt_gc_plot <- plot_genome_wide_rate_bars(
  donor_wt_gc_summary,
  x_col = "donor_wt_gc_bin",
  x_label = "WT donor GC",
  title = "WT donor GC content"
)

safe_ggsave(guide_gc_23_plot, file.path(plots_dir, "guide_gc_23_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(guide_gc_20_plot, file.path(plots_dir, "guide_gc_20_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(guide_gc_18_plot, file.path(plots_dir, "guide_gc_18_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(donor_wt_gc_plot, file.path(plots_dir, "donor_wt_gc_summary.pdf"), width = 5.2, height = 4.8)

gc_dependency_figure <- (
  (guide_gc_23_plot | guide_gc_20_plot) /
  (guide_gc_18_plot | donor_wt_gc_plot)
) +
  patchwork::plot_annotation(tag_levels = "A")

safe_ggsave(
  gc_dependency_figure,
  file.path(plots_dir, "gc_dependency_panels.pdf"),
  width = 11,
  height = 9.5
)

gc_dependency_figure
```

## Colony concordance across repeated barcodes

```{r}
multi_colony_consistency_summary
```

## Non-target variants and SCORE.SV

```{r}
background_variant_burden_plot <- plot_background_variant_burden(background_colony_summary)

safe_ggsave(
  background_variant_burden_plot,
  file.path(plots_dir, "background_variant_burden.pdf"),
  width = 5.5,
  height = 5
)

background_variant_burden_plot
```

```{r}
background_edit_status_summary %>%
  dplyr::mutate(
    mean_high_conf_non_target_count = round(mean_high_conf_non_target_count, 3),
    median_high_conf_non_target_count = round(median_high_conf_non_target_count, 3),
    frac_with_any_high_conf_non_target = scales::percent(frac_with_any_high_conf_non_target, accuracy = 0.1),
    mean_max_non_target_score_sv = round(mean_max_non_target_score_sv, 3)
  )
```

```{r}
recurrent_background_variants
```

## Secondary filters and contextual checks

```{r}
attrition_plot <- plot_genome_wide_stage_attrition(stage_summary)
coverage_qc_plot <- plot_genome_wide_coverage_qc(colony_assays)
max_a_plot <- plot_genome_wide_rate_bars(
  max_a_summary,
  x_col = "max_a",
  x_label = "max A run",
  title = "Long A-runs are at most a secondary filter"
)
max_t_plot <- plot_genome_wide_rate_bars(
  max_t_summary,
  x_col = "max_t",
  x_label = "max T run",
  title = "T-runs were not a clean marginal failure mode"
)
region_class_plot <- plot_genome_wide_rate_bars(
  region_class_summary,
  x_col = "region_class",
  x_label = "Variant region class",
  title = "Locus context affects editability but is not a cassette rule"
)
hotspot_plot <- plot_genome_wide_rate_bars(
  hotspot_summary,
  x_col = "hotspot_flag",
  x_label = "SV hotspot annotation",
  title = "SV hotspots are a contextual caution flag"
)
weighted_model_plot <- plot_genome_wide_weighted_model(weighted_model_table)

safe_ggsave(attrition_plot, file.path(plots_dir, "attrition_plot.pdf"), width = 8, height = 4.5)
safe_ggsave(coverage_qc_plot, file.path(plots_dir, "coverage_qc.pdf"), width = 5.5, height = 5.5)
safe_ggsave(max_a_plot, file.path(plots_dir, "max_a_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(max_t_plot, file.path(plots_dir, "max_t_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(region_class_plot, file.path(plots_dir, "region_class_summary.pdf"), width = 5.6, height = 4.8)
safe_ggsave(hotspot_plot, file.path(plots_dir, "hotspot_summary.pdf"), width = 5.2, height = 4.8)
safe_ggsave(weighted_model_plot, file.path(plots_dir, "weighted_model_odds_ratios.pdf"), width = 8.5, height = 4.8)

supplementary_figure <- (
  (attrition_plot | coverage_qc_plot) /
  (max_a_plot | max_t_plot) /
  (region_class_plot | hotspot_plot)
) +
  patchwork::plot_annotation(tag_levels = "A")

safe_ggsave(
  supplementary_figure,
  file.path(plots_dir, "supplementary_design_rule_context.pdf"),
  width = 11,
  height = 13
)

supplementary_figure
```

## Rule summary

```{r}
design_rule_summary %>%
  dplyr::mutate(
    preferred_design_success_rate = scales::percent(preferred_design_success_rate, accuracy = 0.1),
    other_design_success_rate = scales::percent(other_design_success_rate, accuracy = 0.1),
    preferred_colony_edit_rate = scales::percent(preferred_colony_edit_rate, accuracy = 0.1),
    other_colony_edit_rate = scales::percent(other_colony_edit_rate, accuracy = 0.1)
  )
```

## Weighted model coefficients

```{r}
weighted_model_table %>%
  dplyr::select(
    feature_group,
    term_label,
    odds_ratio,
    odds_ratio_low,
    odds_ratio_high,
    p.value
  ) %>%
  dplyr::mutate(
    odds_ratio = round(odds_ratio, 3),
    odds_ratio_low = round(odds_ratio_low, 3),
    odds_ratio_high = round(odds_ratio_high, 3),
    p.value = signif(p.value, 3)
  )
```

## Extended model coefficients

```{r}
extended_weighted_model_table %>%
  dplyr::select(
    feature_group,
    term_label,
    odds_ratio,
    odds_ratio_low,
    odds_ratio_high,
    p.value
  ) %>%
  dplyr::mutate(
    odds_ratio = round(odds_ratio, 3),
    odds_ratio_low = round(odds_ratio_low, 3),
    odds_ratio_high = round(odds_ratio_high, 3),
    p.value = signif(p.value, 3)
  )
```
